---
title: "A short R heatmap tutorial"
author: "A.Bailey"
date: "29 July 2015"
output: html_document
---
This is a short tutorial for producing heatmaps in R using a test data set 
provided by Leanne Wickens. I assume the reader is able to install packages,
load libraries etc. in RStudio to use this code.

### Loading and preprocessing the data

The global settings for the R markdown file are to show the R code, cache the 
data and load the required packages:
```{r setoptions, echo=TRUE,results='asis',cache=TRUE}
knitr::opts_chunk$set(cache=TRUE,echo=TRUE,cache=TRUE, message=FALSE, 
                      warning=FALSE, fig=TRUE,eval=TRUE,
                      fig.width=9, fig.height = 9,fig.lp='fig:')
library(d3heatmap)
library(gplots)
```

For test data to load, I first provide the URL to Leanne's test data which
contains observations for 63 proteins for three control experiments and three
experiments where the proteins are exposed to TGF-$\beta$.

```{r}
fileUrl <- "https://github.com/ab604/heatmap/blob/master/leanne_testdata.csv"
```

If the data file `leanne_testdata.csv` is not already downloaded, this is 
the code to get it. It also produces a log file:

```{r}        
if(!file.exists("leanne_testdata.csv")){
        download.file(fileUrl,destfile="leanne_testdata.csv")
        dateDownloaded <- date()
        log_con <- file("testdata_download.log")
        cat (fileUrl,"\n","destfile= leanne_testdata.csv",
             "\n","destdir =", getwd(),"\n",dateDownloaded, 
             file = log_con)
        close(log_con)
}
```

Next we load the data and tidy it up a bit. Here first I create some meaningful
column names as `cols` for the accession codes for the proteins in the rows, the three controls and three TGF-$\beta$ experiments respectively. 
These names are used when reading in the `.csv` file with `read.csv` to the `dat` data frame, by declaring `col.names = cols`. `row.names = 1` makes it read column
one, the protein accession codes, as row names.

Here I also set my colour palette as blue, white and red for the heatmap for 
later as `my_palette` with 25 increments.

I then extract columns 1 to 6 to a new data frame called `dat.tdy` as we don't
need the last column. (Try `View(dat)` to see why.)

We normalise the data using `scale`, transposing `dat.tdy` using `t` to 
perform row-wise normalision so that: $z = \frac{x - \mu}\sigma$ to create `dat.n`

Additionally, I've transposed the data back to the original form, but this is
purely for aesthetics.

```{r, preprocess-data}
# Create columnn names for data
cols <- c("Acc","None.1","None.2","None.3","TGFB.1",
          "TGFB.2","TGFB.3","Ttest")

# Set colours for heatmap, 25 increments
my_palette <- colorRampPalette(c("blue","white","red"))(n = 25)

# Load data, re-naming columns and taking column 1 for row names.
dat <- read.csv("leanne_testdata.csv", row.names = 1,
                stringsAsFactors = FALSE, col.names = cols, header = TRUE)

# Tidy the data, extract only columns 1 to 6
dat.tdy <- dat[,1:6]

# Normalize tidy data, transpose for row wise normalisation
dat.n <- scale(t(dat.tdy))

# Put data back in original form for heatmap
dat.tn <- t(dat.n)
```

To check the normalisation, here we calculate the means and standard deviation
for each row (each protein):

```{r, check-norm}
# Check means are zero and std devs are 1
round(colMeans(dat.n),1)
apply(dat.n,2,sd)
```

Next we do the hierarchical clustering needed for the heatmap. First we calculate 
the distance between the rows in `dat.n`. In this case this is linear distance is 
between the experiments for each protein. The experiments are in the rows and the
proteins in the columns in `dat.n`. For example the distance between `None.1` and
`None.2` is `sqrt(sum((dat.n[1,]-dat.n[2,])^2))`. This means we take each
row 2 value from row 1 value, square that result, sum all those results to a 
single value and take the square root to find the linear distance between these
rows, which is `r round(sqrt(sum((dat.n[1,]-dat.n[2,])^2)),3)` and you can check 
this  against the first value in `d1`.

```{r,distance-1}
# Calculate distance between data in columns
d1 <- dist(dat.n,method = "euclidean", diag = FALSE, upper = FALSE)
```

Then we cluster according to these distances.

```{r, cluster-1}
## Column clustering (adjust here distance/linkage methods to what you need!)
c1 <- hclust(d1, method = "ward.D", members = NULL)
```

Then we do the same
```{r, distance-cluster-2}
# Calculate distance between data in rows
d2 <- dist(dat.tn,method = "euclidean", diag = FALSE, upper = TRUE)
# Row clustering (adjust here distance/linkage methods to what you need!)
c2 <- hclust(d2, method = "ward.D", members = NULL)
```

To check this is doing something sensible, lets look at the dendrograms made
clustering our distance matrices `d1` and `d2`:

```{r, dendrogam}
# Check col and row clustering by plotting dendrograms
par(mfrow=c(2,1),cex=0.5) # Shrink labels
plot(c1); plot(c2)

```

```{r, heatmap-1}
## Plot heatmap wit heatmap.2
#pdf(file="test.pdf",paper = "a4")
par(cex.main=0.75) # Shrink title fonts
heatmap.2(dat.tn, Colv=as.dendrogram(c1), Rowv=as.dendrogram(c2), 
          density.info="histogram", trace="none",col = my_palette,
          cexRow=0.5,cexCol=0.75)
#dev.off()
```

```{r, interactive-heatmap,cache=FALSE}
# Create an interactive heatmap with d3heatmap
d3heatmap(dat.tn,Rowv=as.dendrogram(c2),Colv=as.dendrogram(c1),col=my_palette)
```